---
title: "Soil organic matter decomposition rates at agricultural sites in Sweden"
author: "Teneille Nel"
format: html
editor: visual
---

## Overview

Teneille's full project repository is available at <https://github.com/Teneille97/nel_sweden_agsoil_2026>.

## Simple two-pool models

We fit two-pool models using observed C stocks and \Delta $^{14}$C measurements of bulk soil and organic matter fractions, where fractions stable at 325 and 400 degrees are treated as the "fast" and slow-cycling pools, respectively. 

**Key points**
 
-  Because the 325-stable pool has technically already lost the faster-cycling pool that was initially combusted, we see over-predictions of the fractions to compensate for this "missing carbon".
-  The observed \Delta $^{14}$C of bulk soil remains constant despite the predicted increase that was predicted during the atmospheric spike in bomb carbon; this means that very little new carbon was incorporated into soil organic matter, potentially due to erosion and/ or management practices such as tillage or residue removals. 

```{r}
#| echo: false
library(here)

files <- list.files(here("plots/simple_twopool"), pattern = "\\.rds$", full.names = TRUE)

plots <- lapply(files, readRDS)

for (i in seq_along(plots)) {
  print(plots[[i]])
}
```

## Nested two-pool models

We fit two-pool models treating the 325-stable pool, as "bulk" soil and using observed  C stocks and \Delta $^{14}$C measurements of the 325-stable pool and bulk soil. In this nested scenario, the observed 'fast' pool carbon stocks are calculated as the difference between 325-stable and 400-stable carbon stocks, while the fast pool \Delta $^{14}$C values are strictly modeled. 

**Key points**

-  The models show that the observed carbon stocks and \Delta $^{14}$C values of the 325-stable pool are well explained by a heterogeneous system of a faster cycling pool along with the residual pool. 
-  This means that our 325 pool, although modeled initially as a single fast-cycling pool, actually represents a nested pool of intermediate- and slow-cycling carbon.
-  Preliminary attempts to model the intermediate fraction in a 3-pool system did not give good fits. 

```{r}
#| echo: false
library(here)

files <- list.files(here("plots/325asbulk"), pattern = "\\.rds$", full.names = TRUE)

plots <- lapply(files, readRDS)

for (i in seq_along(plots)) {
  print(plots[[i]])
}
```
