---
title: "Soil organic matter decomposition rates at agricultural sites in Sweden - revised"
author: "Teneille Nel"
format: html
editor: visual
---

Teneille's project repository is available on Github at [this link](https://github.com/Teneille97/nel_sweden_agsoil_2026).

## Overview

Four agricultural sites were established in Sweden (M2, M4, M6 and R94) and soils were sampled from 1957 - 2019. Bulk soil samples were fractionated by thermal combustion, isolating two pools (325-\deg C and 400-\deg C stable). Carbon content and radiocarbon signatures of the samples were measured, mean soil bulk density was determined and plant-derived carbon inputs were estimated.

## Four-pool model - parameter optimization

We grouped data by year and sample type (bulk and thermal stability fractions) and calculated averages and uncertainties for carbon stocks and radiocarbon. We calculated carbon stocks in the fast pool by subtracting the 300-$\deg$C-stable pool from bulk soil, and carbon stocks in the intermediate pool by subtracting the 400-$\deg$C-stable pool from the 300-$\deg$C-stable pool. 

The SoilR  package includes functions that help us simulate how carbon moves between pools and decomposes over time. We built a soil carbon model that splits carbon into four pools: fast, intermediate, slow, and carbon lost by some process (potentially erosion). We gave the model: 

• Constant carbon inputs estimated from literature values 

• Observed bulk and organic fraction carbon stocks 

• Bulk soil radiocarbon measurements 

• Atmospheric radiocarbon data

Then we fitted the model to the data using optimization by a cost function (FME), which finds the best parameters (e.g. decomposition rates and transfer between pools) so that the model matches observed  carbon stocks and radiocarbon values. We plotted carbon stocks and radiocarbon of the bulk soil and pools over time to show how well our model predicts these values. Thereafter, we used a Monte Carlo Markov Chain (MCMC) method (from FME) to explore many possible parameter combinations and plot density curves to show likely parameter values. We also checked correlations between parameters and trace plots to make sure the algorithm converged.

**Key points**

The loss pool explains the lack of accumulation of modern atmospheric radiocarbon in the bulk soil despite the fast pool not decomposing so rapidly

```{r}
#| echo: false
#| warning: false
#| message: false

library(here)

readRDS("C:/Users/tnel/OneDrive - Universiteit Antwerpen/Documents/nel_sweden_agsoil_2026/plots/allsites/4_pool/convergence_plot.rds")
readRDS("C:/Users/tnel/OneDrive - Universiteit Antwerpen/Documents/nel_sweden_agsoil_2026/plots/allsites/4_pool/pairs_plot.rds")
readRDS("C:/Users/tnel/OneDrive - Universiteit Antwerpen/Documents/nel_sweden_agsoil_2026/plots/allsites/4_pool/density_plot.rds")

load(here::here("mod_runs/MCMC_allsites_4pools.Rdata"))
mcmc_bestpars<-MCMC$bestpar

param_names_results_4pool <- c("kf", "ki", "ks", "alpha21", "F0ib", "alpha32", "alpha41", "alpha42")
mcmc_pars_df<-data.frame("Par" = param_names_results_4pool, "Value" = mcmc_bestpars)
print(mcmc_pars_df)
```

## Modeled radiocarbon and carbon stocks

We used samples of MCMC parameter sets to run the model many times, which let us estimate uncertainty in model predictions.

**Key points** We found low model uncertainty for radiocarbon and carbon stocks.

```{r}
#| echo: false
#| warning: false
#| message: false

library(here)
C_final <- readRDS("C:/Users/tnel/OneDrive - Universiteit Antwerpen/Documents/nel_sweden_agsoil_2026/plots/allsites/4_pool/C_final.rds")
C14_final <- readRDS("C:/Users/tnel/OneDrive - Universiteit Antwerpen/Documents/nel_sweden_agsoil_2026/plots/allsites/4_pool/14C_final.rds")

print(C_final)
print(C14_final)

```

## Ages and transit times

We used our model to estimate how long carbon stays in the soil system and in each pool (fast, intermediate, and slow). We focused only on these three pools and ignored the “loss” pool, because that pool represents carbon leaving the system permanently. To capture uncertainty, we took many different parameter combinations from our MCMC results (after removing the unstable early part of the chain) and ran the model for each one. For every run, we calculated mean system age, mean pool ages, and mean system transit time. We then combined all runs to calculate the range of uncertainty.

**Key points**

Fast-cycling carbon and overall system transit time are very well constrained, while slower pools have much wider uncertainty.

```{r}
#| echo: false
#| warning: false
#| message: false

library(here)

plot_log_age_tt<-readRDS("C:/Users/tnel/OneDrive - Universiteit Antwerpen/Documents/nel_sweden_agsoil_2026/plots/allsites/4_pool/plot_log_age_tt.rds")
print(plot_log_age_tt)

load(here::here("mod_runs/summary_table_fmt.Rdata"))
print(summary_table_fmt)

```
