---
title: "Soil organic matter decomposition rates at agricultural sites in Sweden"
author: "Teneille Nel"
format: html
editor: visual
---

Teneille's project repository is available on Github at [this link](https://github.com/Teneille97/nel_sweden_agsoil_2026).

## Overview

Four agricultural sites were established in Sweden (M2, M4, M6 and R94) and soils were samples from 1957 - 2019. 
Bulk soil samples were fractionated by thermal combustion, isolating two pools (325-\deg C and 400-\deg C stable). 
Carbon content and radiocarbon signatures of the samples were measured, mean soil bulk density was determined plant-derived carbon inputs were estimated.

## Simple two-pool models

We fit two-pool models (script available [here](https://github.com/Teneille97/nel_sweden_agsoil_2026/blob/main/scripts/14C_models_sweden.R)) using observed C stocks and \Delta $^{14}$C measurements of bulk soil and organic matter fractions, where fractions stable at 325 and 400 degrees are treated as the "fast" and slow-cycling pools, respectively.

**Key points**

-   Because the 325-stable pool has technically already lost the faster-cycling pool that was initially combusted, we see over-predictions of the fractions to compensate for this "missing carbon".
-   The observed \Delta $^{14}$C of bulk soil remains constant despite the predicted increase that was predicted during the atmospheric spike in bomb carbon; this means that very little new carbon was incorporated into soil organic matter, potentially due to erosion and/ or management practices such as tillage or residue removals.

```{r}
#| echo: false
#| warning: false
#| message: false

library(here)

files <- list.files(here("plots/simple_twopool"), pattern = "\\.rds$", full.names = TRUE)

plots <- lapply(files, readRDS)

for (i in seq_along(plots)) {
  print(plots[[i]])
}
```

## Modeling the invisible intermediate pool

We fit two-pool models by ignoring the bulk measurements and creating a system where the 325-stable pool is seen as as "bulk" soil and the 400-stable soil remains the slow pool. In this scenario, the observed 'fast' pool carbon stocks are calculated as the difference between 325-stable and 400-stable carbon stocks, while the fast pool \Delta $^{14}$C values are strictly modeled.

**Key points**

-   The models show that the observed carbon stocks and \Delta $^{14}$C values of the 325-stable pool are well explained by a heterogeneous system of a faster cycling pool along with the residual pool.
-   This means that our 325 pool, although modeled initially as a single fast-cycling pool, actually represents a nested pool of intermediate- and slow-cycling carbon.
-   Preliminary attempts to model the intermediate fraction in a 3-pool nested system did not give good fits.

```{r}
#| echo: false
#| warning: false
#| message: false

library(here)

files <- list.files(here("plots/325asbulk"), pattern = "\\.rds$", full.names = TRUE)

plots <- lapply(files, readRDS)

for (i in seq_along(plots)) {
  print(plots[[i]])
}
```

## Modeling the invisible fast pool

We fit two-pool models with the actual bulk pool and treating the 325-stable pool, as the "slow" pool, leaving out the 400-stable pool. In this nested scenario, the observed fast pool carbon stocks are calculated as the difference between bulk and 325-stable carbon stocks, while the fast pool \Delta $^{14}$C values are strictly modeled.

**Key points**

-   The models show that the observed carbon stocks and \Delta $^{14}$C values of the bulk soil are well explained by a heterogeneous system of a fast-cycling pool along with the 325-stable pool.
-   This means that combustion at 325 degrees removed a fast-cycling fraction which was not captured in our initial models, which treat the 325-pool as a fast fraction, even though it is technically a nested pool of intermediate- and slow-cycling carbon.

```{r}
#| echo: false
#| warning: false
#| message: false

library(here)

files <- list.files(here("plots/325asslow"), pattern = "\\.rds$", full.names = TRUE)

plots <- lapply(files, readRDS)

for (i in seq_along(plots)) {
  print(plots[[i]])
}
```

## Another nested model attempt

We already mentioned that preliminary attempts to model the intermediate fraction in a 3-pool system did not give good fits.
So here We attempted to fit three-pool models with the actual bulk pool and and slow (400-stable) pools, using the modeled 'invisible fast pool'\Delta $^{14}$C values from the previous step, as if they were observed values. In this scenario, the observed intermediate pool carbon stocks are calculated as the difference between 325-stable carbon and 400-stable carbon stocks, while the intermediate pool \Delta $^{14}$C values are strictly modeled.

**Key points**

-   The \Delta $^{14}$C trends of intermediate pool differed a lot between treatments, potentially compensating for bad fits
-  The intermediate pool, although small, seems to play an important role in explaining the observed bulk \Delta $^{14}$C values
-  In some treatments, the pool is highly stable, but in one treatment it incorporates an unreasonable amount of modern carbon from the fast pool - perhaps the systems collapses from an initial 3-pool system to a 2-pool system due to changes management at some point in time?
-  Perhaps knowledge about historic management practices and Bayesian uncertainty distributions can help to narrow down the likely ranges of transfer

```{r}
#| echo: false
#| warning: false
#| message: false

library(here)

files <- list.files(here("plots/325asslow/3_pool"), pattern = "\\.rds$", full.names = TRUE)

plots <- lapply(files, readRDS)

for (i in seq_along(plots)) {
  print(plots[[i]])
}
```
